---
title: "Difference-in-Differences Tutorial"
author: "Ohi Islam"
date: "`r Sys.Date()`"
output:
  github_document: default
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
---

# Introduction

When introducing the difference-in-differences (DiD) method in Economics of Education, it is often helpful to begin with intuition before turning to regression. This tutorial follows a step-by-step approach. We first examine outcomes visually, then organize them in a simple table, and finally connect that setup to a regression model. The goal is to help students understand how the DiD estimate captures the additional change in outcomes for the treated group relative to the control group.

# Learning Goals

By the end of this tutorial, students should be able to:

- describe the logic of a DiD design
- compute a DiD estimate using average outcomes
- interpret the interaction term in a regression model
- connect graphical, tabular, and regression-based interpretations


##A Simple Example

Suppose we are studying the effect of an education policy introduced for one group of schools but not another. We observe average outcomes for treatment and control groups before and after the policy.

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
```

# Visual Intuition

Suppose one group is exposed to a policy intervention and another group is not. We observe outcomes for both groups before and after the intervention.

```{r create-data}
did_df <- tibble(
  group = c("Control", "Control", "Treatment", "Treatment"),
  period = c("Pre", "Post", "Pre", "Post"),
  outcome = c(50, 54, 48, 60)
)

did_df
```

We can visualize these outcomes first.

```{r plot-bar}
ggplot(did_df, aes(x = period, y = outcome, fill = group)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Outcomes by Group and Period",
    x = "Period",
    y = "Average Outcome"
  )
```

The graph allows us to compare changes over time for the treatment and control groups. The control group improves somewhat between the pre-treatment and post-treatment periods, while the treatment group improves by more. The key question is whether the additional improvement in the treatment group can be attributed to the policy.


# Constructing the DiD Estimate

The DiD estimate compares:

1. the change in the treatment group
2. the change in the control group
3. the difference between those two changes

## Step 1: Compute Within-Group Changes
Now compute the change over time within each group.


```{r manual-did}
control_pre  <- 50
control_post <- 54
treat_pre    <- 48
treat_post   <- 60

control_change <- control_post - control_pre
treat_change   <- treat_post - treat_pre
did_estimate   <- treat_change - control_change

control_change
treat_change
did_estimate
```

The control group changed by `r control_change`, and the treatment group changed by `r treat_change`.

So the DiD estimate is:

\[
(60 - 48) - (54 - 50) = 12 - 4 = 8
\]

## Step 2: Show the Calculation in a Table
We now simulate individual-level data so we can estimate the same idea using regression.


```{r did-table}
did_summary <- tibble(
  Group = c("Control", "Treatment"),
  Pre = c(control_pre, treat_pre),
  Post = c(control_post, treat_post),
  Change = c(control_change, treat_change)
)

kable(did_summary, caption = "Outcomes and Changes by Group")
```

This table shows that the treatment group improved more than the control group, and the extra improvement is the DiD estimate.

# Regression Interpretation

We now connect this idea to a regression model.

\[
Y_i = \beta_0 + \beta_1 \text{Treatment}_i + \beta_2 \text{Post}_i + \beta_3 (\text{Treatment}_i \times \text{Post}_i) + \varepsilon_i
\]

The coefficient \(\beta_3\) is the DiD estimate.

## Simulated Individual-Level Data

To fit the regression, we create a simple individual-level dataset that matches the same group averages.

```{r simulate-data}
set.seed(123)

n <- 100

sim_df <- tibble(
  treatment = rep(c(0, 0, 1, 1), each = n),
  post = rep(c(0, 1, 0, 1), each = n),
  mean_outcome = c(
    rep(50, n),
    rep(54, n),
    rep(48, n),
    rep(60, n)
  )
) %>%
  mutate(
    outcome = rnorm(n = 4 * n, mean = mean_outcome, sd = 4)
  )

head(sim_df)
```

## Estimate the DiD Regression

```{r run-regression}
did_model <- lm(outcome ~ treatment + post + treatment:post, data = sim_df)

summary(did_model)
```

# Interpreting the Coefficients

In this regression:

- the intercept is the average outcome for the control group in the pre period
- the coefficient on `treatment` is the difference between treatment and control in the pre period
- the coefficient on `post` is the change over time for the control group
- the coefficient on `treatment:post` is the DiD estimate

## Extract the Key Coefficient

```{r extract-beta3}
coef(did_model)["treatment:post"]
```

This interaction coefficient represents the additional post-period change in the treated group relative to the control group.

# Linking the Graph, Table, and Regression

The same treatment effect appears in three ways:

- in the graph, as the larger increase for the treatment group
- in the table, as the difference between the two group-specific changes
- in the regression, as the interaction coefficient

This is why the DiD framework is useful pedagogically: students can see the same idea visually, numerically, and econometrically.

## Why the Regression Coefficients Are Not Exactly Equal to the Table Differences

In the earlier table, the outcomes were specified as deterministic group
averages.

  Group       Pre   Post   Change
  ----------- ----- ------ --------
  Control     50    54     4
  Treatment   48    60     12

From these values we can compute the theoretical differences:

-   Pre-treatment difference: 48 b 50 = b2\
-   Change in the control group: 54 b 50 = 4\
-   Change in the treatment group: 60 b 48 = 12

The difference-in-differences estimate is therefore

(60 b 48) b (54 b 50) = 12 b 4 = 8.

These numbers come directly from the table and are therefore exact.

However, in the regression section we did not use deterministic
outcomes. Instead, we simulated individual-level data with random
variation around the group means:




# Parallel Trends Assumption

For a DiD design to have a causal interpretation, we usually rely on the parallel trends assumption.

This means that, in the absence of treatment, the treatment and control groups would have followed similar trends over time.

This assumption cannot be tested directly for the treated post-treatment outcome, but researchers often examine pre-treatment trends and institutional context to assess whether it is plausible.

# Conclusion

Difference-in-differences estimates the treatment effect by comparing changes over time across treated and untreated groups. The key parameter is the interaction between treatment status and the post-treatment period.

# Optional Exercise

Try modifying the four group-period averages and see how the DiD estimate changes.

For example:

- increase the post-treatment outcome for the treatment group
- decrease the post-treatment outcome for the control group
- make both groups trend upward by the same amount


How does each change affect the DiD estimate?
